### 2023-01-22 doc/response.tex

We recognize that our writing was unclear before.
But we know from the observation that the large outbreak during the Omicron period was driven by superspreading event related to a campus event.
We could not describe the event in detail due to privacy reasons, which led to unclear writing. 
We have tried to make the causality clearer throughout (see response to Reviewer 1's comment 9).
We also account for the possibility of shorter infectious periods and lower vaccine effectiveness.
See below for detailed responses for each point.

\revtext{Main comments:}

\revtext{Fig. 1- you might consider converting case counts for Princeton and Mercer County into cases per 100,000 per day (or some other denominator quantity). These plots are comparing cases for populations of ~15,000 for Princeton and ~360,000 for Mercer County, which makes Fig. 1 panels I-L a bit difficult to parse at first—by controlling the denominator you could more easily compare to a 1:1 slope for panels E-J and a 1:1 ratio for panels I-L. This approach would also account for the changing student body size at Princeton across semesters with reopening, etc.}

Thank you for the suggestion. 
We have changed the main figure and corresponding supplementary figures for New York and Philadelphia to show weekly cases per 100,000.
We have revised the writing accordingly and added a new Supplementary Figure comparing cases per 1000 for each population group on Princeton campus (Figure S1).
We decided to keep the raw cases for comparisons with model projections since we are no longer comparing populations of two different sizes.

\revtext{Lines 165-169- a key piece of this story seems to be variant predominance—you could consider adding a new figure with variant proportions through time or dashed vertical lines to Fig. 1 time courses indicating when a new variant reached dominance.}

We have variant data but we feel analyzing variant data is beyond the scope of this paper.
We decided not to add the variant data in this paper also because it involves a different group of individuals analysing the data.
And so we prefer to keep that analysis as a separate paper.

\revtext{Lines 214-226: the assumption that testing patterns remain the same on campus and off campus throughout the entire investigation period doesn’t seem realistic. In particular for the beginning of the Omicron period, the Princeton university campus continued with surveillance screening, while the general population likely experienced significantly decreased ascertainment rates in part due to the increased availability and use of at home tests which are not mandated reporting (e.g., one estimate of a change of ascertainment from 50\% pre-Omicron to 20-29\% during BA.1; \url{https://www.medrxiv.org/content/10.1101/2022.04.22.22274198v3.full.pdf}). Therefore, I would hesitate to conclude that the lack of correlation was indicative of a lack of community transmission—changes in testing patterns could really be driving that pattern for the final semester.}

Thank you for pointing this out. 
We meant that the testing patterns remained constant for each semester, rather than the entire investigation period.
We also meant that the case ratio would stay the same if testing patterns and transmission conditions were to remain constant, but we are not necessarily assuming that the testing patterns remained constant.
We have tried to make the writing of this paragraph clearer while acknowledging changes in testing patterns over the investigation period.

Also, we know that the outbreak during the final semester was associated with a specific campus event among students on one specific day, which we unfortunately cannot described in detail due to privacy reasons. 
So we know that the lack of correlation was likely driven by a lack of community transmission in this case. 
We have tried to clarified this point.

\revtext{Lines 243-245- Infectious periods changed with variants, e.g., Omicron << Delta. Was that component explored at all?}

We now provide simulations that assume shorter infectious and latent period during the Omicron outbreak:

``We considered the possibility that the Omicron variant can have shorter latent and infectious periods by decreasing the mean duration of latent, pre-symptomatic, and (a)symptomatic stages of infection by 0.5 days (therefore a total of 1.5 reduction in the duration of infection).
In this case, a shorter generation interval can lead to faster growth rate given the same values of $\Rc$ \citep{wallinga2007generation}.
However, we find that the effects of shorter infection has small effects on the overall dynamics (Supplementary Figure S13).''

\revtext{Lines 296-305- While reading through this the first time, I was wondering about differences in vaccine effectiveness during Omicron emergence however—maybe hint to the reader here that you address that later or separately?}

We have added the following sentence:

``We note that these assumptions are specific to the Delta variant---we discuss vaccine effectiveness against the Omicron variant later on.''

\revtext{Lines 453-454- I think your vaccine effectiveness estimates against Omicron are still quite optimistic, which could explain why ramping up your R-value didn’t really do much. In practice, we’ve seen closer to 30-50\% effectiveness against infection for example vs. 70\% pre-waning (\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9398552/})}

We now provide simulations that assume 30\% vaccine effectiveness as a supplementary figure with the following text in the main manuscript:

``We also considered the possibility that the vaccine effectiveness against the Omicron variant might be lower by repeating the same analysis with 30\% effectiveness against infection \citep{tan2023vaccine}.
When the baseline $\Rc$ is low ($\Rc=2$), increasing $\Rc$ still does not increase the number of cases sufficiently.
When we assume an intermediate value of $\Rc=4$, the model does a better job at capturing the dynamics but it does so by overestimating the trough before the policy change and underestimating the peak after the policy change.
When we assume a high value of $\Rc=6$, the model overestimates both the trough and the peak.''

\revtext{Fig. 3-4- Apologies if you’ve stated this elsewhere and I’ve missed it, but have you taken a look at your testing volume through time? I know that the PU policy was x \# of tests per week based on vaccination status, but how was that enforced? I’m wondering if there is as much variation in the y-axis if you look at test positivity vs. raw number of cases? Another thing to consider, since I’m not sure about testing cadence and frequency would be to do some rolling averages to help smooth out differences across days/weeks.}

We now show testing volume through time in Supplementary Figure S1.
Everyone was required to be tested in order to be present on campus. 
And as we see in Supplementary Figure S1, there was very good compliance and stable testing volumes within each semester, except during the end of fall semester of the 2021--2022 academic year (reflecting Omicron circulations) and the beginning of spring semester of the 2021--2022 academic year (reflecting students returning to school).
Even then, these changes do not correlate with increase in cases---for example, the increase in testing volume at the end of fall semester of the 2021--2022 academic year is very sudden (doubling in a week), but the changes in cases at the same period is much smaller.

We agree there is variation in cases within a week but testing patterns were fairly stable across weeks, which is why we chose to look at weekly data, instead of daily data.
We feel that showing raw variation in weekly cases (rather than rolling averages) would be more accurate because we are also able to identify the sources of most variations (e.g., school holidays and classes ending).

\revtext{Lines 456-457- instead of artificially introducing new infections, what happens if you decrease $k$ your overdispersion parameter to increase the potential variance (i.e., allow for bigger gatherings?)}

Artificially introducing new infection is meant to emulate the increase in heterogeneity.
Decreasing $k$ alone would certainly increase variance but we have no control over when large outbreaks will happen, meaning that most of the simulations will not be able to match the data.

\revtext{Minor comments:}

\revtext{Line 40- double check consistency of variant name capitalization throughout (e.g., “omicron” vs. “Omicron”)}

Fixed.

\revtext{Lines 78-81- infrastructure including ventilation would also play a key role here}

We have added ventilation infrastructure to this sentence.

\revtext{Line 192- “counties” instead of “countries”?}

Fixed.

\revtext{Lines 270-271- in reality, those contacts are likely much more structured leading to a more modular network (e.g. dormmates would have repeated contacts, students would be more likely to have classes with others in that same major); how would you see less mixing affecting your model results and interpretation?}

We have added the following paragraph:

``In reality, the contact structure among the campus population is likely more structured, exhibiting strong assortativity.
For example, undergraduate students are more likely to mix with other undergraduate students, rather than graduate students or faculty and staff members.
Even among undergraduate students, students are more likely to mix with their close friend group than with other students.
On one hand, assortative mixing may lead to faster epidemic growth within certain population groups;
on the other hand, it can also make the disease more difficult to spread among other groups that have lower contact rates.
Therefore, predicting the impact of structured contact network requires more detailed information about whether the majority of cases were infected at random or from certain groups of the campus population.
For simplicity, we assume random mixing throughout the paper---nonetheless, allowing for overdispersion in transmission is expected to emulate variability in epidemic growth rates driven by complex contact structures \citep{lloyd2005superspreading}.''

\revtext{Line 278- it would be helpful to label $\theta$ as your “community transmission” parameter on first use, since you refer to it that way later, but not stated explicitly here.}

Based on suggestions by Reviewer 1, we now normalize $\theta$ by population size $N$ and provide a clear definition the first time it is used>

\revtext{Line 387- “cannot” instead of “can”?}

``can'' is correct. We have revised this sentence to make the point clearer:

``These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time)---we see that extending the simulation beyond November 26th still captures the increase in cases.''

\revtext{Lines 162-164- state date of policy implementation to make link clearer for Figs. 3 \& 4}

Done.

\bibliography{university-covid}

\end{document}
